### 2024-02-27 thesis/sections/6-grounding-language-games.tex
thesis

\section{Grounding referring expressions in language games}
\label{sec:language-games}
The experiments that are executed using language games have a similar structure as the experiments in the previous chapters, since those provided the basis for the language games.
% SD: But there is an importnat difference that now we have two neural netweorks, the sender and the receiver.
% DK: TODO

@@

All agents are trained using the \emph{Adam} optimizer \citep{Kingma2015}.
LSTMs are used to produce and parse the message between the agents.
Gumbel-Softmax relaxation is applied to train the agents with those discrete messages.
Hereby, a temperature of $\tau=1$ is used which showed the best results in preliminary experiments.

>>>>

\section{Grounding referring expressions in language games}
\label{sec:language-games}
The experiments that are executed using language games have a similar structure as the experiments in the previous chapters, since those provided the basis for the language games.
However, every task is solved by two neural networks that communicate through a discrete bottleneck.
% SD: But there is an importnat difference that now we have two neural netweorks, the sender and the receiver.
% DK: TODO

@@

Across all experiments, the same restriction on the vocabulary and the produced messages are compared.
First, five different vocabulary sizes $|V|$ are tested: 2, 10, 16, 50 and 100 symbols.
Notice that these vocabularies include the \emph{<eos>} token which signals the end of the message.
A vocabulary of two symbols therefore consists effectively of only one token that can be learned by the agents.
The center of attention lies on the vocabularies with 16 and 50 symbols.
The vocabulary with 16 symbols corresponds the 13 attributes the objects can have and align with human language referring expressions.
However, a smaller vocabulary size increases the pressure to converge on meaning, which is my also 10 and 2 symbols are tested.
On the other side, the agents might instead of a compositional language learn to assign one symbol with every combination of attributes.
This results in $2 * 3 * 8 = 42$ unique combinations, which corresponds to the vocabulary with 50 symbols.
Furthermore, a vocabulary of 100 symbols is used to test the effect of a much bigger bottleneck.

Secondly, also different maximum message lengths are compared: 1, 2, 3, 4 and 6 symbols.
Notice that this applies only to the maximum possible length; the agents can choose to use shorter messages.
Again, the corresponding length to human referring expressions in English is 3 for the three attributes.
However, the agents can be pressured to condense the information more precisely with shorter messages or given less restrictions with longer messages.
Vocabulary size and message length play hereby an opposing role in the pressure on condensing information.
% If the symbols were similarly used, messages would have lengths between one and three symbols.
% Opposed to that a slightly smaller vocabulary with 10 symbols is used to create a smaller bottleneck with a higher pressure to condense the information.
% Similarly, a bigger vocabulary consisting of 20 symbols tests how a bigger bottleneck changes the results.
% Lastly, a big vocabulary of 100 symbols should give the model all options to encode the information, including one symbol per attribute or one symbol describing a combination of attributes.
% SD: Motivation why this is so. The possible number of strings is {#colour} x {#shape} {object type} where all numbers except the last one also contain a padding token.
% DK: TODO

All agents are trained using the \emph{Adam} optimizer \citep{Kingma2015}.
LSTMs are used to produce and parse the message between the agents.
Gumbel-Softmax relaxation is applied to train the agents with those discrete messages.
Hereby, a temperature of $\tau=1$ is used which showed the best results in preliminary experiments.
All setups are additionally compared to a baseline.
In the baseline the sender is "blinded", meaning that the sender is not shown any input but produces random messages.
Effectively, the receiver needs to solve the task on their own, by exploiting underlying structures in the shown data and the target data.
If a non-baseline model doesn't achieve a better performance than the baseline, the sender didn't communicate any meaningful messages and no language emerged.

### 2024-02-27 thesis/sections/6.1-object-identification.tex
thesis

\subsection{Object identification}
\subsubsection*{Setup}

@@

The discrimination games in this research have a very similar setup as described in \citep{Lazaridou2017}.
The agents in this research resemble their \emph{agnostic sender} as well as their \emph{receiver}.
One central difference is the production of the message.
The main goal of their language games was the identification of the concept that the shown image was related to.
Therefore, the sender communicated only single-symbol messages to the receiver which should describe the concept of the target image.
Opposed to that, in this research, the agents are tasked to discriminate objects from each other based on their attributes.
% SD: On the contrary,...
% DK: (done)
It is therefore assumed that the sender will communicate these discriminative attributes.
For that reason, the sender is allowed to generate sequences as a message where for instance each symbol in the sequence might correspond to on attribute.

The data is prepared in the same way as for the single model object identification task in section \ref{sec:object-identification}.
Bounding boxes of all objects in the scene are extracted and preprocessed to be used for the feature extractors.
Accordingly, also the models from the single model tasks serve as basis for the object identification task.
The sender uses the architecture of the model described in section \ref{sec:referring_expression_generation} apart from the final LSTM.
Instead, it will use the LSTM with the hidden size $h_s$ included in the \emph{EGG} framework to produce a message.
Based on the results from the single model object identifiers, the image embedding size $e_s$ is fixed to 100.
The bounding box of the target object is passed first, the remaining are shuffled.

The receiver instead uses the architecture of the single model object identifier, described in section \ref{sec:object-identification}.
Similarly to the sender, the LSTM of the \emph{EGG} framework with the hidden size $h_e$ is used to encode the sender's message instead of the one-hot encoded attributes in the single model.
The hidden state of the LSTM is projected to the embedding size $e_r=500$ with a linear layer, based on the results in the previous experiments.
As before the dot product between the embedded message and the encoded image is calculated to let the model point towards one of the bounding boxes.
Opposed to the sender, the receiver is shown all bounding boxes in shuffled order.
Figure \ref{fig:discriminator_architecture} shows, how the sender and the receiver of the discriminator are built up.

@@

The experiments are conducted with a learning rate of $2\times10^{-4}$, the loss is calculated using cross entropy.
The following values for the variables are compared:
\begin{itemize}
    \item $h_s$: 100, 500, 1000
    \item $h_r$: 10, 30, 50, 100, 500, 1000
    \item $|V|$: 1, 10, 16, 50, 100
    \item $n$: 1, 2, 3, 4, 5, 6
\end{itemize}
The language game is run on the 'Dale-2', 'Dale-5' and 'CLEVR color' datasets.

\subsubsection*{Results}
Table \ref{tab:results_discriminator} shows the accuracy of the models calculated on the success of communication if the receiver can identify the target object.
A random guess corresponds to 50\% in the \emph{Dale-2} dataset and 20\% in the \emph{Dale-5} dataset.
Four different vocabulary sizes $|V|$ are tested.
A size of 13 symbols corresponds to the 13 attributes the objects can have and align with human language.
If the symbols were similarly used, messages would have lengths between one and three symbols.
Opposed to that a slightly smaller vocabulary with 10 symbols is used to create a smaller bottleneck with a higher pressure to condense the information.
Similarly, a bigger vocabulary consisting of 20 symbols tests how a bigger bottleneck changes the results.
Lastly, a big vocabulary of 100 symbols should give the model all options to encode the information, including one symbol per attribute or one symbol describing a combination of attributes.
% SD: Motivation why this is so. The possible number of strings is {#colour} x {#shape} {object type} where all numbers except the last one also contain a padding token.
% DK: TODO

>>>>

\subsection{Object identification}
\label{sec:language-games:identification}

\subsubsection*{Setup}

@@

The discrimination games in this research have a very similar setup as described in \citep{Lazaridou2017}.
The agents in this research resemble their \emph{agnostic sender} as well as their \emph{receiver}.
One central difference is the production of the message.
The main goal of their language games was the identification of the concept that the shown image was related to.
Therefore, the sender communicated only single-symbol messages to the receiver which should describe the concept of the target image.
Opposed to that, in this research, the agents are tasked to discriminate objects from each other based on their attributes.
% SD: On the contrary,...
% DK: (done)
It is therefore assumed that the sender will communicate these discriminative attributes.
For that reason, the sender is allowed to generate sequences as a message where for instance each symbol in the sequence might correspond to one attribute.

The data is prepared in the same way as for the single model object identification task in section \ref{sec:object-identification}.
Bounding boxes of all objects in the scene are extracted and preprocessed to be used for the feature extractors.
For datasets with a variable number of objects, the list bounding of bounding boxes is padded with a matrix of zeros to the maximum possible number of objects present in a scene across the
dataset.
Opposed to the experiments with one neural model, no additional information about the target object's attributes is provided.
Instead, the sender agent is presented with the target object always as the first object, while the distractors are shuffled.
The receiver is presented with completely shuffled bounding boxes.

Accordingly, also the models from the single model tasks serve as basis for the object identification task.
The sender uses the architecture of the model described in section \ref{sec:referring_expression_generation} apart from the final LSTM.
Instead, it will use the LSTM with a hidden size $h_s$ included in the \emph{EGG} framework to produce a message.
Based on the results from the single model object identifiers, the image embedding size $e_s$ is fixed to 100.

The receiver instead uses the architecture of the single model object identifier, described in section \ref{sec:object-identification}.
Similarly to the sender, the LSTM of the \emph{EGG} framework with the hidden size $h_e$ is used to encode the sender's message instead of the one-hot encoded attributes in the single model.
The hidden state of the LSTM is projected to the embedding size $e_r=500$ with a linear layer, based on the results in the previous experiments.
As before the dot product between the embedded message and the encoded image is calculated to let the model point towards the bounding box that is described best by the message.
Figure \ref{fig:discriminator_architecture} shows, how the sender and the receiver of the discriminator are built up.

@@

The experiments are conducted with a learning rate of $2\times10^{-4}$, the loss is calculated using cross entropy.
The following values for the variables are compared:
\begin{itemize}
    \item $h_s$: 100, 500, 1000
    \item $h_r$: 10, 30, 50, 100, 500, 1000
    \item $|V|$: 2, 10, 16, 50, 100
    \item $n$: 1, 2, 3, 4, 6
\end{itemize}

The language game is run on the 'Dale-2', 'Dale-5' and 'CLEVR color' datasets.
A random guess corresponds to 50\% in the \emph{Dale-2} dataset, 20\% in the \emph{Dale-5} dataset and 13\% in the \emph{CLEVR color} dataset.

\subsubsection*{Results}
When looking at the results of the experiments, it can be seen that the agents are successfully discriminating the objects across all datasets.
Table TODO gives an overview over the accuracy scores of selected configurations.
For each dataset there are many configurations that outperform the baseline models by a high margin.
However, there are still clear differences between the accuracy scores of each dataset.

When the agents are tasked to discriminate between the two objects of the \emph{Dale-2} dataset, it succeeds in almost all configurations with an accuracy of over 94\%.
In many configurations the accuracy lies above 99\%, while only 5 configurations have an accuracy of only 61\% to 64\%.
An indicator for

Table \ref{tab:results_discriminator} shows the accuracy of the models calculated on the success of communication if the receiver can identify the target object.

### 2024-02-27 thesis/sections/6.2-re-generation.tex
thesis

In a next step, it is tested if the agents can learn to extract features of the objects together.
For this, the receiver is tasked to describe the target object in natural language, while the sender needs to communicate, which object is the target.
Again, the setup is asymmetrical: the sender receives the image and information, which of the objects in the image is the target object in form of a masked image.
The receiver only sees the image without additional information.
\cmtDK[inline]{human bias vs emergent language}

For this experiment, the data is prepared in the same way as described in section \ref{sec:referring_expression_generation}.
Since the position of the padding as well as the order of the words didn't have an effect on the results, the usual approach in natural generation task is used: padding tokens are appended and the referring expression is not reversed.
Respectively, both agents are based on the models in these experiments.
The sender uses the \emph{masked RE generator} with its LSTM replaced with the LSTM included in the \emph{EGG} framework, using the hidden size $h_s$.
\cmtDK[inline]{fixed parameters}
The receiver is built up similarly.
The image is encoded in the same way, but afterwards it is concatenated with the decoded message, received be the sender with the hidden size $h_r$.
The concatenated is reduced to $d=$ dimensions, based on the results of the previous experiments, and used as the initial state of the captioning LSTM.
\cmtDK[inline]{fixed parameters}

@@

During training, the ground truth caption is used as the input to the LSTM using teacher forcing.
% SD: I'm not sure how teacher forcing is implemented, see my earlier comment
% DK: typo, teacher forcing is actually applied (done)
When presented with test data, the LSTM always produces three tokens, by using its own predicted words as the input for the next step.

The experiments are conducted with the following hyperparameters: a learning rate of $2\times10^{-4}$, a temperature for the Gumbel-Softmax relaxation of 1 and \emph{Adam} \citep{Kingma2015} as optimizer.
The loss is calculated using cross entropy.
The following values for the variables are compared:
\begin{itemize}
    \item $h_s$: 100, 500, 1000
    \item $h_r$: 10, 30, 50, 100, 500, 1000
    \item $|V|$: 1, 10, 16, 50, 100
    \item $n$: 1, 2, 3, 4, 5, 6
\end{itemize}
\cmtDK[inline]{real values}
Since the agents are trained to describe the target object discriminatively based on the described GRE-algorithm, they are trained on the 'Dale-2' and 'Dale-5' dataset.
The 'Dale-5' should be again much harder to learn, since there are more objects that the agents need to discriminate the target object from.
% SD: So the sender is generating descriptions following the GRE policy?
% DK: TODO
The same metrics as in the section \ref{sec:referring_expression_generation} are used to evaluate the results.

>>>>

In a next step, it is tested if the agents can learn to extract features of the objects together.
For this the same sender as in the previous section is used.
Based on the results, $h_s$ is fixed to $500$ and $e_s$ to $100$
However, the task and by that the receiver's architecture is adapted in two different ways.
In the first setup, the \textbf{referring expression generator}, the receiver is tasked to describe the target object with natural language following the \emph{incremental GRE-algorithm} based on the sender's message (see Figure \ref{fig:caption_generator_game_architecture}).
This follows the approach of the approach of the \emph{RE generators} described in section \ref{sec:referring_expression_generation}.
The whole scene is encoded using the \emph{image encoder} submodule, projecting it to the image encoding dimension $e_{ri}=100$.
The sender's message is decoded with the hidden size $h_r=500$ and $e_r=100$, and concatenated with the encoded image.
This is then reduced to $LSTM_o=1500$ dimensions based on the results of the previous experiments, and used as the initial state of the captioning LSTM.
Tokens are embedded with $LSTM_e=15$ dimensions.
Since the position of the padding as well as the order of the words in the target didn't have an effect on the results, the usual approach in natural generation task is used: padding tokens are appended and the referring expression is not reversed.
During training, the ground truth caption is used as the input to the LSTM using teacher forcing.
% SD: I'm not sure how teacher forcing is implemented, see my earlier comment
% DK: typo, teacher forcing is actually applied (done)
When presented with test data, the LSTM always produces three tokens, by using its own predicted words as the input for the next step.
The loss is calculated using cross entropy.

@@

However, this setup could lead to two problems:
First, the task is quite complex for both agents to learn, since two language generation steps are involved.
This might stop the agents from converging towards a useful language.
Secondly, in case a language emerges, it could be aligned with the natural language of the target referring expressions.
In other words, the sender might just learn to effectively repeat the target referring expressions without extracting the necessary features themselves.
For that reason, in the second setup the receiver is tasked to predict only the attributes of the target object in the shape of a one-hot vector.
By this, no natural language including its salience order is involved and the complexity of the complete game is reduced.
The receiver uses the same way to encode and combine the complete scene and the sender's message with $e_{ri}=100$, $h_r=500$ and $e_r=100$.
This is then passed to a linear layer that produces a vector with 13 dimension, corresponding to the 2 sizes, 3 shapes, and 8 colors.
The loss is then calculated using binary cross entropy.

\cmtDK[inline]{figure}

The experiments for both setups are conducted with a learning rate of $2\times10^{-4}$.
As in the previous section, the following values for the variables are compared:
\begin{itemize}
    \item $|V|$: 2, 10, 16, 50, 100
    \item $n$: 1, 2, 3, 4, 6
\end{itemize}

Since the agents are trained to describe the target object discriminatively based on the described GRE-algorithm, they are trained on the \emph{Dale-2}, \emph{Dale-5} and \emph{CLEVR color} dataset.
The \emph{Dale-5} and the \emph{CLEVR color} should be again much harder to learn, since there are more objects that the agents need to discriminate the target object from.
% SD: So the sender is generating descriptions following the GRE policy?
% DK: TODO
The same metrics as in the section \ref{sec:referring_expression_generation} are used to evaluate the results for the first setup.
When predicting one-hot vectors, an overall \textbf{accuracy} is reported that reflects if all attributes were predicted correctly.
Additionally, the accuracy for each attribute is calculated separately to show which attributes give bigger challenges for the models.

### 2024-02-27 thesis/sections/6.3-reference-resolution.tex
thesis

\subsection{Reference resolution}
\subsubsection*{Setup}
In the final experiments, the agents are tasked to produced and understand referring expressions only based on the implicit human knowledge present in the scenes as for example the attributes of the objects.
Opposed to the previous experiments, explicit human language information as human referring expressions or one-hot encoded attributes are avoided.
By this, the agents can't just reuse these human referring expressions, but need to learn to generate them themselves during the language games just based on the visual input.
This is done similarly to the single model reference resolution task, described in section \ref{sec:reference_resolution}.
The agents are tasked to predict the center coordinates of the target object, after being presented with the complete scene.
However, the agents, more specifically the sender needs to be pointed towards the target object.
In the final setup, the \emph{masked reference resolution game}, this is done providing the sender with an additional masked visual input.
However, this can prove to be a challenging task for the agents.
For this reason, in an approach to reduce the complexity, for the \emph{RE reference resolution game} the sender is shown a human referring expression, generated with the incremental GRE algorithm instead of the additional masked image.
This might lead to a heavy bias of the emerged language towards the human referring expressions and will be compared to the unbiased emerged language.

The masked sender is the same as the \emph{RE generation sender}, described in section \ref{sec:referring-expression-generation-game}.
The RE sender works similarly, but instead encodes the human referring expression with an LSTM.
The final hidden state is concatenated with the encoded image.
The resulting vector is reduced to the hidden size $h_s$ and used as the initial hidden state of the LSTM included in the \emph{EGG} framework to generate the message.
\cmtDK[inline]{fixed variables}

The receiver is the same in both setups and is based on the \emph{reference resolver} + referring expressions, described in section \ref{sec:reference_resolution}.
Again, instead of the included LSTM, the receiver uses the \emph{EGG} LSTM with the hidden size $h_r$ to encode the message.
The dimensions of the coordinate predictor are fixed to $c=$, based on the previous results.
\cmtDK[inline]{fixed variables}
% SD: Just thinking: would there be a more systematic way to compare all the models? Perhaps if you write the parameters in a table or a text that is parallel. Then you could simply say, modal A + these changes. It would be much clearer to the reader how all models are rerlated and it would be easier for the reader to follow.
% DK: TODO
The euclidean distance between the resulting prediction of the center point and the true center of the target object is calculated and the weights of both agents are adapted accordingly.

@@

The experiments are conducted with the following hyperparameters: a learning rate of $2\times10^{-4}$, a temperature for the Gumbel-Softmax relaxation of 1 and \emph{Adam} \citep{Kingma2015} as optimizer.
The loss is calculated using cross entropy.
The following values for the variables are compared:
\begin{itemize}
    \item $h_s$: 100, 500, 1000
    \item $h_r$: 10, 30, 50, 100, 500, 1000
    \item $|V|$: 1, 10, 16, 50, 100
    \item $n$: 1, 2, 3, 4, 5, 6
\end{itemize}
\cmtDK[inline]{real values}

As in section \ref{sec:reference_resolution}, the agents are trained first on the 'CLEVR single' dataset to understand if they are capable of predicting locations in an image together.
In a next step, the 'Dale-2' and 'Dale-5' datasets are used to test if the agents are able to first communicate a target object and second describe the target object discriminatively with a small vocabulary.
The same metrics are used to evaluate the results.

>>>>

\subsection{Reference resolution}
\subsubsection*{Setup}
In the final experiments, the agents are tasked to produced and understand referring expressions only based on the implicit human knowledge present in the scenes, as for example the attributes of the objects.
Opposed to the previous experiments, explicit human language information as human referring expressions or one-hot encoded attributes are avoided.
By this, the agents can't just reuse these human referring expressions, but need to learn to generate them themselves during the language games just based on the visual input.
This is done similarly to the single model reference resolution task, described in section \ref{sec:reference_resolution}.
As before, the sender has the same architecture as in the previous language games, encoding the bounding boxes of all objects.

The receiver is tasked to solve two tasks.
First, in the \textbf{coordinate reference resolver}, the agents are tasked to predict the center coordinates of the target object (see Figure \ref{fig:coordinate_predictor_game_architecture}).
The receiver is the same in both setups and is based on the \emph{reference resolver} + referring expressions, described in section \ref{sec:reference_resolution}.
Again, instead of the included LSTM, the receiver uses the \emph{EGG} LSTM with the hidden size $h_r=500$ and $e_r=100$ to encode the message.
The dimensions of the coordinate predictor are fixed to $c=2048$, based on the previous results.
% SD: Just thinking: would there be a more systematic way to compare all the models? Perhaps if you write the parameters in a table or a text that is parallel. Then you could simply say, modal A + these changes. It would be much clearer to the reader how all models are rerlated and it would be easier for the reader to follow.
% DK: TODO
The euclidean distance between the resulting prediction of the center point and the true center of the target object is calculated and the weights of both agents are adapted accordingly.

@@

Secondly, in the \textbf{attention reference resolver}, the agents learn to point towards regions in the image, which is based on the \emph{attention reference resolver} described in section \ref{sec:reference_resolution}.
As before, the LSTM is replaced with the \emph{EGG} LSTM to parse the sender's message with the hidden size $h_r=500$ and $e_r=100$.
The dot product is calculated between each encoded region of the image and the encoded message, and the $softmax$ function is applied subsequently.
The loss is calculated using \emph{binary cross entropy} between the predictions and ground truth regions.

\cmtDK[inline]{figure}

The experiments are conducted with the following hyperparameters: a learning rate of $2\times10^{-4}$, a temperature for the Gumbel-Softmax relaxation of 1 and \emph{Adam} \citep{Kingma2015} as optimizer.
The following values for the variables are compared:
\begin{itemize}
    \item $|V|$: 2, 10, 16, 50, 100
    \item $n$: 1, 2, 3, 4,  6
\end{itemize}

As in section \ref{sec:reference_resolution}, the agents are trained first on the \emph{CLEVR single} dataset to understand if they are capable of predicting locations in an image together.
In a next step, the \emph{Dale-2}, \emph{Dale-5} and \emph{CLEVR color} datasets are used to test if the agents are able to first communicate a target object and second describe the target object discriminatively with a small vocabulary.
The same metrics are used to evaluate the results as described in \ref{sec:reference_resolution}.
